Red States = Red Flags?

### 2024-02-15 paper/paper.qmd
complete paper

---
title: "Red States = Red Flags?"
subtitle: "An analysis of birth rate trends from 2001 to 2019 and an exploration of political affiliation of different states and Birth rates in the United States,"
author: 
  - Rayan Awad Alim
  - Maria Mangru
  - Isha Junega
thanks: "Code and data are available at: https://github.com/RayanAlim/Birth_Rate_Analysis"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
#bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

# Workspace Set up
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(readr)
library(maps)
library(haven)
library(readxl)
library(here)
library(lmtest)
library(sandwich)
```

Introduction

@@

Melissa S. Kearney, Phillip B. Levine, and Luke Pardue, analyze
potential reasons for a decline in birth rate in their paper, “The
Puzzle of Falling US Birth Rates since the Great Recession.” In their
analysis, examine birth rates by demographic groups defined by age,
education, race and ethnicity, marital status, and birth parity inorder
to find clues that serve as an explanation for the decline. They found
factors pointing to the time and location do not serve as enough
explanation to the decline in birthrates, although less quantifiable,
women shifting their priorities to their aspirations for life instead of
having more children is an important contributor. Their analysis does
not explore solid evidence of US-specific policies or economic factors
that can explain the depth of that decline(). We address this gap by
replicating their paper and focusing on wages, access to healthcare
(with religious importance) and childcare expenditure data to draw
conclusions about whether these factors affect the decline or birth
rates. We found that XX. This is important because. This paper is
structured to include the data, results, and the discussion that include
our analysis. Although the code used in the original paper is Strata, we
use R to replicate some of its graphs, as well as to conduct analysis of
our own.

Data

Melissa S. Kearney, Phillip B. Levine, and Luke Pardue's paper "The
Puzzle of Falling US Birth Rates since the Great Recession" provided one
of the datasets used in this analysis, notably the policyvars01_19
dataset. This dataset presents a thorough summary of state-level
policies that may have an impact on birth rates in the US from 2001 to
2019 by assembling data on numerous policy variables from numerous
sources.

Article:

Model

@@

Model set-up
=======
Where:

@@

Results

1- Medicaid coverage has increased which allows more low-income people
to get an abortion.

1- Medicaid coverage has increased which allows more low-income people
to get an abortion.

2- Maximum well-fare has gone down so it more difficult to raise a kid

3- Parents seem to have more of an interest in having an abortion which
aligns with the decrease in number of births

2- Maximum well-fare has gone down so it more difficult to raise a kid

3- Parents seem to have more of an interest in having an abortion which
aligns with the decrease in number of births

Analysis Results

### Analysis Results

# State-Specific Birth Averages (2001-2019):

@@

```{r echo=FALSE, message=FALSE}

@@

# Create the bar plot with scaled averages
ggplot(average_births_by_state, aes(x = reorder(stname, -AverageBirths), y = AverageBirths)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Scaled Average Number of Births by State (2001-2019)",
       x = "State",
       y = "Scaled Average Number of Births") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Child Expenditure

Figure C2 The number of people with ACA Medicaid increased dramatically
from 2000 to about 2014. Then, since 2014, it has slowly increased
non-decreasingly till 2020.

Figure C3 There has been a linear decrease of the maximum welfare
benefits from 2000 to 2015. From 2015 to 2020 there has been a small
rise in the maximum welfare benefits. Overall, there has been a loss in
the maximum welfare benefits, despite the small rise.

Figure C4 The blue portion of each column represents the number of
parents who have consent for an abortion. The red portion represents the
number of parents who don’t have consent for an abortion. There is a
slight increase in consent for abortion from 2001 to 2019

# State Minimum Wage Changes (2001-2019):

Figure 5 shows the trend of the average minimum wage across states
depicted by the blue line and the average number of births represented
by the red line, between the years 2001 to 2009. The trend for average
minimum wage shows a noticeable increase over time. In contrast, the
average number of births shows a fluctuating trend. Starting from 2001,
the number of births gradually increase, with a peak occurring around
2007. After this peak, there is a discernible decline in the average
number of births.

```{r echo=FALSE, message=FALSE, results='asis'}

@@

# Plot the graph with adjusted scales
ggplot(average_data) +
  geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", size = 1) +
  geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", size = 1) +
  scale_y_continuous(
    "Minimum Wage",
    sec.axis = sec_axis(~ . / scaling_factor, name = "Number of Births")
  ) +
  theme_minimal() +
  labs(title = "Comparison of Average Minimum Wage and Birth Rates (2001-2009)",
       x = "Year")

@@

# Impact of Abortion Delay Policy

@@

```{r echo=FALSE, message=FALSE, results='asis'}
# Define file paths 
policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
```

The range of birth rates in 2001 spans from approximately 17 per 1000
(excluding Utah (UT)) people in certain states to below 15 per 1000 in
others, with a convergence of rates around 12 to 14 per 1000.

The states in this graph are ones which had a policy change from no
delay to delay which allows for the analysis of the before-and-after
effects of policy implementation.

For example, Arizona's (AZ) implementation year of 2009 shows birth rate
descends sharply afterward. In contrast, North Carolina (NC) in 2011
didn't show such an immediate shift.

The states without delay policies showcase a natural decline in birth
rates over time, without the potential influence of policy changes. This
trend is consistent across the states, with some variations in the
steepness of the decline.

For example, California (CA) maintains a relatively high birth rate over
the years, starting near 15 per 1000 in 2001 and ending around 12 per
1000 in 2019. This could be influenced by its socio-economic profile and
immigration patterns.

All states show a general decline in relative birth rates over the
19-year period, which is in line with the national trend of declining
birth rates.

State-Specific Trends:

California (CA): The trend in California shows a relatively steady
decline throughout the period, without any sharp changes.

Florida (FL): Florida's trend is similar to California's, with a gradual
decline.

North Carolina (NC): North Carolina, with a policy change in 201, shows
a relatively stable trend before the policy change and a sharper decline
after.

Texas (TX): Texas exhibited a noticeable decline after the policy
implementation year of 2003.

When contrasting the birth rates of Texas and California, two states
with large populations, we see that Texas's birth rates declined more
sharply after its policy change in 2003 compared to California. This
might suggest an association between the policy implementation and the
accelerated decline in birth rates in Texas.

Comparing Georgia and North Carolina, both of which implemented delay
policies. Georgia's birth rates were rising before the policy change and
then started declining after, while North Carolina's birth rates were
relatively stable before 2011 and then declined.

# Education

```{r echo=FALSE, message=FALSE}
#Education

# Creating the plot
ggplot(educ_data, aes(x = year)) +
  geom_line(aes(y = numbirth_lesshs, color = "Less than HS"), linewidth = 1) +
  geom_line(aes(y = numbirth_hsgrad, color = "HS Graduate"), linewidth = 1) +
  geom_line(aes(y = numbirth_somecoll, color = "Some College"), linewidth = 1) +
  geom_line(aes(y = numbirth_colgrad, color = "College Graduate"), linewidth = 1) +
  geom_line(
    aes(y = numbirthalleduc, color = "All Education Levels"),
    size = 1,
    linetype = "dashed"
  ) +
  scale_color_manual(
    values = c(
      "Less than HS" = "blue",
      "HS Graduate" = "green",
      "Some College" = "orange",
      "College Graduate" = "red",
      "All Education Levels" = "grey"
      )
  ) +
  labs(
    title = "Number of Births by Education Level Over Time",
    x = "Year",
    y = "Number of Births",
    color = "Education Level"
  ) +
  theme_minimal() +
  theme(legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

# State Political Affiliation and Birth Rates
```{r echo=FALSE, message=FALSE}
# https://stackoverflow.com/questions/29381652/plotting-us-heat-map-with-text

Discussion

## First discussion point: Abortion

#relative average \# of births in states delay

```{r echo=FALSE, message=FALSE}
# Determine which states had a delay or no delay policy 
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%

@@

Across the 19-year span, the data for states which have not implemented
an abortion policy showcases a wide range of trends. Several states
demonstrate a peak in birth rates at various points during the study
period. While there is variation between the states, the overall
trajectory indicated that the early to late 2000s were a period of
higher birth rates, with a shift toward lower birth rates post 2010. It
is important to note that this data in Figure 7 showcases the relative
average number of births per 1000 people and fluctuates from 10 to
approximately 16.

```{r echo=FALSE, message=FALSE, results='asis'}
# Define file paths 
policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"

@@

In contrast, figure \[\] looks at the trend of relative average number
of births which have consistently implemented a delay policy. Similar to
states which have not implemented a delay policy, birth rates for these
states have generally decreased over the 19 year period. Over this
period, the relative average number of births per 1000 people fluctuated
significantly, ranging from around 10 to over 20.

```{r echo=FALSE, message=FALSE, results='asis'}

@@

Lastly, figure \[\] looks at states which transitioned from having no
abortion delay to policy to implementing this policy at various years
(indicated by the broken lines). A common trend among these states was
an initial period of relatively higher birth rates in the early 2000s,
with subsequent declines over the years. This trend also closely follows
what observed for states which implemented an abortion delay policy and
those which have not implemented such a policy.

# Difference-in-Difference

```{r echo=FALSE, message=FALSE, results='asis'}
# Define file paths 
policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"

@@

# Education

#Politicial Affiliation

### Discussion

# Medicaid - remove heading after The Affordable Care
(ACA)’s Medicaid is “intended to address systemic health inequalities
for millions of Americans who lack health insurance” Medicaid is
“expected to provide coverage to low-income individuals,” and also
covers low-income pregnant women. Excluding the states where abortion is
banned, Medicaid covers abortions, making it more accessible. According
to Figure C1, More people continue to have medicaid as seen based on the
trend from 2000 to 2020, making an affordable abortion more accessible.
In Figure C3, we can see that there is an increase in parental consent
for abortion. In other words, more parents are increasingly interested
in getting an abortion every year. A potential motivator for parents to
want an abortion is that as a low-income family, it is harder to take
care of a larger family. Welfare is “federal government-sponsored
assistance programs for individuals and families in need.” In Figure C2,
it is evident that there is a decrease in maximum welfare benefits,
making it less affordable for lower-income families to have a larger
family size.The increasing demand for an abortion and the insurance
covering it may contribute to a decline in birthrates.

# Abortion policy - remove heading after

Through the difference-in-difference analysis, the results indicate a
decrease of approximately 0.777 births per 1000 people in North Carolina
post policy implementation relative to TN, which did not implement the
abortion delay policy. This decrease is statistically significant,
suggesting that implementing an abortion delay policy may have a real
effect on birth rates.

@@

Second discussion: Political affiliation and birth rates

@@

Fourth discussion point: Education

```{r}
#Education graph
```

Weaknesses and next steps A weakness of this paper is a lack of
economical context. For example, we analyze our data independent of
events like the Recession that happened in 2008. Without providing
context to current events in the time period, our paper gives a weaker
justification for the decline in birthrates. Next steps for improving
our analysis include not only providing context for the events happening
in the US, but focusing on the birth rates on a more micro level. Since
each state within the US has its own laws, it will provide more insight
to focus on the birth rates within each state. Additionally, to improve
our analysis furthermore, we can compare the US to other countries that
are similar from an economical standpoint and compare their birth rate
to the US.

Weaknesses and next steps should also be included.

@@

\appendix

Appendix

Additional data details

Model details

```{r}

>>>>

---
title: "Red States = Red Flags?"
subtitle: "An analysis of birth rate trends from 2001 to 2019 and an exploration of political affiliation of different states and Birth rates in the United States,"
author: 
  - Rayan Awad Alim
  - Maria Mangru
  - Isha Junega
thanks: "Code and data are available at: https://github.com/RayanAlim/Birth_Rate_Analysis"
date: today
date-format: long
abstract: "Melissa S. Kearney, Phillip B. Levine, and Luke Pardue in their paper, 
'The Puzzle of Falling US Birth Rates since the Great Recession', analyze the 
decreasing birth rates within the US. In this paper, we repicate use their data 
and replicate their results of their study. While they focused on examining 
birth rates by demographics, we focus on minimum wages, education, abortion 
poilicy, and child support enforcement. The importance of our research is that 
it highlights the underlying cause of why parents may not to have many children, 
which leads to a decline in birthrates."
format: pdf
number-sections: true
#bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

# Workspace Set up
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(readr)
library(maps)
library(haven)
library(readxl)
library(here)
library(lmtest)
library(sandwich)

# manages graph placement
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
```

# Introduction {#sec-intro}

@@

Melissa S. Kearney, Phillip B. Levine, and Luke Pardue, analyze
potential reasons for a decline in birth rate in their paper, “The
Puzzle of Falling US Birth Rates since the Great Recession.” In their
analysis, examine birth rates by demographic groups defined by age,
education, race and ethnicity, marital status, and birth parity in order
to find clues that serve as an explanation for the decline. They found
factors pointing to the time and location do not serve as enough
explanation to the decline in birthrates, although less quantifiable,
women shifting their priorities to their aspirations for life instead of
having more children is an important contributor. Their analysis does
not explore solid evidence of US-specific policies or economic factors
that can explain the depth of that decline. We address this gap by
replicating their paper and focusing on wages, abortion policies,
education and childcare support enforcement data to draw conclusions
about whether these factors affect the decline or birth rates.

In this paper, we created a linear regression model to determine the
impact of abortion policy on number of birthrates. We analyzed the
affects of education on the number of birthrates by tracking the the
birthrate for each education level separately. Additionally, to gage the
impact of childcare support enforcement on the number of birthrates, we
looked at how the maximum welfare benefits, including Medicaid, affect
low-income individuals. Analyzing the wages, abortion policies,
education, and childcare support important because they give deeper
insights regarding why parents may not prefer to have many children and
therefore may cause a decline in birthrates. The paper is structured to
include the data, results, and the discussion that include our analysis.
Although the code used in the original paper is Strata, we use R to
replicate some of its graphs, as well as to conduct analysis of our own.

# Data {#sec-data}

## Data Source and Methodology

As part of our study, we looked closely at the variables influencing the decline in US birth rates. Our goal was to expand on and close the gaps left by Melissa S. Kearney, Phillip B. Levine, and Luke Pardue in their paper, "The Puzzle of Falling US Birth Rates since the Great Recession." The interaction between birth rates and childcare costs, healthcare availability, education, political affiliation, and incomes was the main focus of our research. We replicated many of the visuals and analyses from the original paper using R for data visualization and analysis, building on its conclusions through the use of a variety of datasets and approaches.
The primary datasets for our study came from a variety of sources, such as the UK Center for Poverty Research National Welfare Database, the annual policy analysis data, and comprehensive birth records from 2001 to 2019. The Center for Poverty Research National Welfare Database [@ukcprWelfareData] at the University of Kentucky supplied more data on state-level laws, minimum wage levels, and political affiliations to supplement these original statistics. The UKCPR National Welfare Data which is a comprehensive dataset that has been compiled and annually updated by the University of Kentucky Center for Poverty Research (UKCPR). This database includes many variables such as population, wage levels, political landscape, dynamics of poverty, employment, and welfare in the United States at the state level. Covering a wide time span from 1980 to 2021, this dataset provides a longitudinal view of significant socio-economic indicators across the country.
Building on earlier work by Kearney and Levine [@Kearney2022Puzzle], state-specific abortion delay law data was updated using the Guttmacher Institute's "Counseling and Waiting Periods for Abortion" database. The Kaiser Family Foundation's "Status of State Medicaid Expansion Decisions: Interactive Map" (KFF 2021) and a collection of birth statistics were the sources of information on Medicaid expansions.
The methods used to gather the data made it possible to analyze state-level policies that potentially affect US birth rates in detail between 2001 and 2019. Through the integration of publicly accessible datasets and meticulously curated policy variables, our study offered a comprehensive understanding of the diverse factors influencing the fluctuations in US birth rates.

## Variables and Summary Statisitcs
With an emphasis on the years 2001 to 2019, we created and improved a collection
of variables which are needed for studying the decrease in birth rates 
throughout the United States. Our results are stronger because we were able to 
record the short-term impacts of various different socioeconomic determinants 
on birth rates during this time. The development of important variables, their 
importance, storage formats, and the interactions between them are described in 
detail below. Graphs, summary statistics, and comments are used to provide 
context.

The first main variable of interest is the delay variable which looks at the 
Abortion Policy Delay. This binary variable indicated whether or not a stated 
has implemented a delay in abortion policy. Our Difference-in-Differences (DiD) 
analysis relied on this variable to evaluate the policy's effect on birth rates.
Secondly relative_births was created to determine the relative birth rates of 
each state by population in order to provide a common metric for comparing birth
rates among states with different population sizes. This variable was created 
by combining birth statistics with population estimates. State Minimum Wage 
(min_wage) indicates the minimum wage for each state in a certain year. 
The purpose of including this economic variable in the study of the correlation
between birth rates and salary levels was to test the hypothesis that family
planning choices could be influenced by greater salaries. Healthcare Access 
(healthcare_access): An integer-based composite index based on the availability 
of healthcare facilities and the expansion of Medicaid. This variable sought 
to measure the ease of access to healthcare services, as this could have an 
effect on birth rates by affecting the availability of prenatal and postnatal 
care. Next, the variable childcare_exp quantified whether or not parents 
received any support for their childcare expenses. Considering the substantial 
expenses linked to childrearing, this variable was essential to comprehending 
the financial influences on decisions about family size.

High-level cleaning included resolving missing numbers, guaranteeing consistency
across datasets, and filtering data to the years 2001-2019. Harmonizing 
variables such as state_name made proper merging possible. To aid in analysis, 
numerical variables were inspected for anomalies and, if required, normalized.

Visual insights into the effects of abortion policy delays and economic factors
on birth rates were provided via graphs showing the trend in relative birth 
rates over time, by state, and by policy status. A summary of the distribution 
and dispersion of the data was given via summary statistics. For example, the 
distribution of min_wage brought attention to economic inequities, while the 
average relative birth rate and its standard deviation showed variances between
states and years.

The associations between the important variables were shown using scatter plots
and correlation analysis. For example, the relationship between min salary and 
relative births was investigated in order to comprehend the impact of the 
economy on family planning choices.

## Measurements

These datasets have several limitations even if they are quite useful for our 
research. Firstly, reporting variations in birth rates can occur between states 
and over time. These disparities could be the result of delayed data entry, 
modifications to reporting requirements, or variations in data collection 
techniques. These discrepancies could distort the birth rate estimates, 
causing an overestimation or underestimating in particular states.

Secondly, although the policy data is extensive, it might not fully account 
for all state-level programs that have the potential to affect birth rates. 
In addition, they do not give insight into variations within the policy. Birth
trends may also be significantly influenced by policies that are not covered by 
the dataset, such as regional healthcare access initiatives or educational 
initiatives, but these factors are not taken into consideration in our research.
To address these limitations, we used several strategies. To confirm the trends 
found and guarantee consistency across sources, we cross-referenced our primary 
data sources with supplementary datasets and research. Next, in order to 
determine the potential impact of data mistakes, we performed sensitivity 
analysis. We evaluated the robustness of our findings to differences in data 
quality by correcting for any under- or over reporting in birth data.

# Results: {#sec-results}

## Model

@@

### Model set-up

Where:

@@

## Birth Rate Results

### State-Specific Birth Averages (2001-2019):

@@

```{r echo=FALSE, message=FALSE}
#| label: fig-1

@@

# Create the bar plot with scaled averages
ggplot(average_births_by_state, aes(x = reorder(stname, -AverageBirths), y = AverageBirths)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Scaled Average Number of Births by State (2001-2019)",
       x = "State",
       y = "Scaled Average Number of Births") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## State Minimum Wage Changes (2001-2019):

Figure 2 shows the trend of the average minimum wage across states
depicted by the blue line and the average number of births represented
by the red line, between the years 2001 to 2009. The trend for average
minimum wage shows a noticeable increase over time. In contrast, the
average number of births shows a fluctuating trend. Starting from 2001,
the number of births gradually increase, with a peak occurring around
2007. After this peak, there is a discernible decline in the average
number of births.

```{r echo=FALSE, message=FALSE, results='asis'}
#| label: fig-2

@@

# Plot the graph with adjusted scales
ggplot(average_data) +
  #geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", size = 1) +
  #geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", size = 1) +
  geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", linewidth = 1) +
  geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", linewidth = 1) +
  scale_y_continuous(
    "Minimum Wage",
    sec.axis = sec_axis(~ . / scaling_factor, name = "Number of Births")
  ) +
  theme_minimal() +
  labs(title = "Comparison of Average Minimum Wage and Birth Rates (2001-2009)",
       x = "Year")

@@

## Impact of Abortion Delay Policy

@@

```{r echo=FALSE, message=FALSE, results='asis'}
#| label: fig-3
# Define file paths 
policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"

@@

Across the 19-year span, the data for states which have not implemented
an abortion policy showcases a wide range of trends. Several states
demonstrate a peak in birth rates at various points during the study
period. While there is variation between the states, the overall
trajectory indicated that the early to late 2000s were a period of
higher birth rates, with a shift toward lower birth rates post 2010. It
is important to note that this data in Figure 3 showcases the relative
average number of births per 1000 people and fluctuates from 10 to
approximately 16.

```{r echo=FALSE, message=FALSE, results='asis'}
#| label: fig-4
# Define file paths 
policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"

@@

In contrast, figure 4 looks at the trend of relative average number
of births which have consistently implemented a delay policy. Similar to
states which have not implemented a delay policy, birth rates for these
states have generally decreased over the 19 year period. Over this
period, the relative average number of births per 1000 people fluctuated
significantly, ranging from around 10 to over 20.

```{r echo=FALSE, message=FALSE, results='asis'}
#| label: fig-5

@@

Lastly, figure 5 looks at states which transitioned from having no
abortion delay to policy to implementing this policy at various years
(indicated by the broken lines). A common trend among these states was
an initial period of relatively higher birth rates in the early 2000s,
with subsequent declines over the years. This trend also closely follows
what observed for states which implemented an abortion delay policy and
those which have not implemented such a policy.

### Difference-in-Difference

```{r echo=FALSE, message=FALSE, results='asis'}
#| label: tbl-1
# Define file paths 
policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"

@@

## Education

```{r echo=FALSE, message=FALSE}
#| label: fig-6

#Education

# Creating the plot
ggplot(educ_data, aes(x = year)) +
  geom_line(aes(y = numbirth_lesshs, color = "Less than HS"), linewidth = 1) +
  geom_line(aes(y = numbirth_hsgrad, color = "HS Graduate"), linewidth = 1) +
  geom_line(aes(y = numbirth_somecoll, color = "Some College"), linewidth = 1) +
  geom_line(aes(y = numbirth_colgrad, color = "College Graduate"), linewidth = 1) +
  geom_line(
    aes(y = numbirthalleduc, color = "All Education Levels"),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_color_manual(
    values = c(
      "Less than HS" = "blue",
      "HS Graduate" = "green",
      "Some College" = "orange",
      "College Graduate" = "red",
      "All Education Levels" = "grey"
      )
  ) +
  labs(
    title = "Number of Births by Education Level Over Time",
    x = "Year",
    y = "Number of Births",
    color = "Education Level"
  ) +
  theme_minimal() +
  theme(legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

TODO: Results description for Education

### State Political Affiliation and Birth Rates

```{r echo=FALSE, message=FALSE}
#| label: fig-7
# https://stackoverflow.com/questions/29381652/plotting-us-heat-map-with-text

TODO: Results description for Political Affiliation

## Child Support Enforcement

```{r echo=FALSE, message=FALSE}
child_enforce_data <- read_dta("../data/raw_data/annual_policy/policyvars01_19.dta")
```

The number of people with ACA Medicaid increased dramatically
from 2000 to about 2014. As figure 8 depicts, since 2014, the number of people 
with Medicaid has slowly increased non-decreasingly till 2020. The ACA Medicaid 
expansion has been adopted and implemented by states at different times, which 
explains some of the dramatic increases within the data.

```{r echo=FALSE, message=FALSE, fig.cap= "Expansion of Medicaid from 2000 to 2020"}
#| label: fig-8
#| 
# medicaid benefits
ggplot(child_enforce_data, aes(x = year, y = aca_mcaid_exp)) +
  geom_col(fill = "#1f77b4" ) + labs(x = "Year", y = "Number of people with ACA Medicaid ", title = "People with ACA Medicaid Over Time (Years)") + theme_minimal() +  theme(plot.title = element_text(hjust = 0.5))  
```

In Figure 9, there has been a linear decrease of the maximum welfare benefits
from 2000 to 2015. From 2015 to 2020 there has been a small rise in the maximum 
welfare benefits. Overall, there has been a loss in the maximum welfare 
benefits, despite the small rise.

```{r echo=FALSE, message=FALSE}
#| label: fig-9
# max well fair benefits
ggplot(child_enforce_data, aes(x=year, y=logben3)) +
    geom_point()  + labs(x = "Year", y = "Max Welfare Benefits", title = "Max Welfare Benefits Over Time (Years)") + theme_minimal() +  theme(plot.title = element_text(hjust = 0.5))  
```

In Figure 10, the blue portion of each column represents the number of
parents who have consent for an abortion. The red portion represents the
number of parents who don’t have consent for an abortion. There is a
slight increase in consent for abortion from 2001 to 2019.

```{r echo=FALSE, message=FALSE}
 #| label: fig-10
# Parental
ggplot(child_enforce_data, aes(x = factor(year), fill = factor(parental))) +
  geom_bar(position = "stack") +
  labs(x = "Year", y = "Parental Consent for Abortion", fill = "Interest", title = "Distribution of Interest in Abortion by Parent Over Time (Years)") + theme_minimal() +  # Apply minimal theme
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, vjust = 0.5))  
```

# Discussion:

## Abortion Policy
Through the difference-in-difference analysis, the results indicate a
decrease of approximately 0.777 births per 1000 people in North Carolina
post policy implementation relative to TN, which did not implement the
abortion delay policy. This decrease is statistically significant,
suggesting that implementing an abortion delay policy may have a real
effect on birth rates.

@@

## Second discussion: Political affiliation and birth rates

@@

## Education

## Child Support Enforcement

(ACA)’s Medicaid is “intended to address systemic health inequalities
for millions of Americans who lack health insurance” Medicaid is
“expected to provide coverage to low-income individuals,” and also
covers low-income pregnant women. Excluding the states where abortion is
banned, Medicaid covers abortions, making it more accessible. According
to Figure C1, More people continue to have medicaid as seen based on the
trend from 2000 to 2020, making an affordable abortion more accessible.
In Figure C3, we can see that there is an increase in parental consent
for abortion. In other words, more parents are increasingly interested
in getting an abortion every year. A potential motivator for parents to
want an abortion is that as a low-income family, it is harder to take
care of a larger family. Welfare is “federal government-sponsored
assistance programs for individuals and families in need.” In Figure C2,
it is evident that there is a decrease in maximum welfare benefits,
making it less affordable for lower-income families to have a larger
family size.The increasing demand for an abortion and the insurance
covering it may contribute to a decline in birthrates.

## 4.2 Weakness and Next Steps

A weakness of this paper is a lack of economical context. For example,
we analyze our data independent of events like the Recession that
happened in 2008. Without providing context to current events in the
time period, our paper gives a weaker justification for the decline in
birthrates. Next steps for improving our analysis include not only
providing context for the events happening in the US, but focusing on
the birth rates on a more micro level. Since each state within the US
has its own laws, it will provide more insight to focus on the birth
rates within each state. Additionally, to improve our analysis
furthermore, we can compare the US to other countries that are similar
from an economical standpoint and compare their birth rate to the US.

@@

# References
